### 2015-03-15 doc/thesis/introduction.tex
Minor improvements to introduction, review and begining of typing rules

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\emph{optional type systems} \citep{bracha2004pluggable} and
\emph{gradual typing} \citep{siek2006gradual} are two similar
approaches for blending static and dynamic typing in the same
language.
The aim of both approaches is to offer programmers the option
to use type annotations where static typing is needed and to avoid
them where dynamic typing is sufficient.
The difference between these two approaches is the way they treat
run-time semantics.
While optional type systems do not affect run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

@@

Typed Lua's intended use is as an application language, and
we believe that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
checked by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications evolve into complex systems rather than just scripts.
We shall see that Typed Lua introduces the refinement of
tables to support the common way that Lua programmers use
them to represent both modules and objects.

@@

In Chapter \ref{chap:review} we review the literature about static
and dynamic typing in the same language, we discuss the differences
between optional type systems and gradual typing, and we also
present the results of our survey on Lua idioms.
In Chapter \ref{chap:typedlua} we use code examples to present the
design of Typed Lua.
In Chapter \ref{chap:system} we use typing rules to present the
formalization of Typed Lua.
In Chapter \ref{chap:evaluation} we discuss some case studies that
we used to evaluate our design, and we also compare our design
to related work.
In Chapter \ref{chap:conc} we outline our contributions.

>>>>

Ideally, programming languages should offer programmers the
option to choose between static and dynamic typing:
\emph{optional type systems} \citep{bracha2004pluggable} and
\emph{gradual typing} \citep{siek2006gradual} are two similar
approaches for blending static and dynamic typing in the same
language.
The aim of both approaches is to offer programmers the option
to use type annotations where static typing is needed,
allowing the incremental migration from dynamic to static typing.
The difference between these two approaches is the way they treat
run-time semantics.
While optional type systems do not affect run-time semantics,
gradual typing uses run-time checks to ensure that dynamically typed
code does not violate the invariants of statically typed code.

@@

Typed Lua's intended use is as an application language, and
we believe that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
checked by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications evolve into complex systems rather than just scripts.
We will show that Typed Lua introduces the refinement of
tables to support the common idioms that Lua programmers use
to encode both modules and objects.

@@

This work is split into seven chapters.
In Chapter \ref{chap:review} we review the literature about blending
static and dynamic typing in the same language, we discuss the differences
between optional type systems and gradual typing, and we also
present the results of our survey on Lua idioms.
In Chapter \ref{chap:typedlua} we use code examples to present the
design of Typed Lua.
In Chapter \ref{chap:system} we present our type system.
In Chapter \ref{chap:evaluation} we discuss the evaluation
results that we obtained while using Typed Lua to type existing Lua code.
In Chapter \ref{chap:related} we present some related work.
In Chapter \ref{chap:conc} we outline our contributions.

### 2015-03-15 doc/thesis/review.tex
Minor improvements to introduction, review and begining of typing rules

Dart \citep{dart} and TypeScript \citep{typescript} are new
languages that are designed with an optional type system.
Both use JavaScript as their code generation target because
their main purpose is web development.
In fact, Dart is a new class-based object-oriented language with
optional type annotations and semantics that resembles the
semantics of Smalltalk, while TypeScript is a strict superset of
JavaScript that provides optional type annotations and class-based
object-oriented programming.
Dart has a nominal type system, while TypeScript has a structural
one, but both are unsound by design.
For instance, Dart has covariant arrays, while TypeScript has
covariant parameter types in function signatures,
besides the interaction between statically and dynamically
typed code that is also unsound.

@@

\begin{figure}[!ht]
\dstart
$$
\begin{array}{llr}
T ::= & & \textsc{types}\\
& \;\; \Number & \textit{base type number}\\
& | \; \String & \textit{base type string}\\
& | \; ? & \textit{dynamic type}\\
& | \; T \rightarrow T & \textit{function types}\\
e ::= & & \textsc{expressions}\\
& \;\; l & \textit{literals}\\
& | \; x & \textit{variables}\\
& | \; \lambda x{:}T{.}e & \textit{abstractions}\\
& | \; e_{1} e_{2} & \textit{application}
\end{array}
$$
\dend
\caption{Syntax of the gradually-typed lambda calculus}
\label{fig:gtlc}
\end{figure}

@@

Figure \ref{fig:consistent-subtyping} combines consistency and subtyping to
compose the consistent-subtyping relation for the simply typed
lambda calculus extended with the dynamic type $?$.
When we combine consistency and subtyping, we are making subtyping handle
what casts are safe among static types, and we are making consistency
handle the casts that involve the dynamic type $?$.
The consistent-subtyping relation is not transitive, and thus
the dynamic type $?$ is not neutral to this relation.

@@

Another concern about casts is how to improve debugging support,
because a cast application can be delayed and the error related
to that cast application can appear considerable distance
from the real error.
\citet{wadler2009wpc} developed \emph{blame calculus} as a way
to handle this issue, and \citet{ahmed2011bfa} extended
blame calculus with polymorphism.
Blame calculus is an intermediate language to integrate
static and dynamic typing that uses the blame tracking approach
proposed by \citet{findler2002chf}.

@@

There are some projects that incorporate gradual typing into some
programming languages.
Reticulated Python \citep{reticulated,vitousek2014deg} is a research
project that evaluates the costs of gradual typing in Python.
Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
that introduces a new cast insertion strategy for gradually-typed
objects \citep{allende2013cis}.
Grace \citep{black2012grace,black2013sg} is a new object-oriented,
gradually-typed, educational language.
In Grace, modules are gradually-typed objects, that is, modules
may have types and methods as attributes, and can have a state
\citep{homer2013modules}.
ActionScript \citep{moock2007as3} is one the first languages that
incorporated gradual typing to its implementation and
Perl 6 \citep{tang2007pri} is also being designed with gradual typing,
though there is little documentation about the gradual type systems
of these languages.

>>>>

Dart \citep{dart} and TypeScript \citep{typescript} are new
languages that are designed with an optional type system.
Both use JavaScript as their code generation target because
their main purpose is Web development.
In fact, Dart is a new class-based object-oriented language with
optional type annotations and semantics that resembles the
semantics of Smalltalk, while TypeScript is a strict superset of
JavaScript that provides optional type annotations and class-based
object-oriented programming.
Dart has a nominal type system, while TypeScript has a structural
one, but both are unsound by design.
For instance, Dart has covariant arrays, while TypeScript has
covariant parameter types in function signatures,
besides the interaction between statically and dynamically
typed code that is also unsound.

@@

\begin{figure}[!ht]
\dstart
$$
\begin{array}{llr}
T ::= & & \textsc{types:}\\
& \;\; \Number & \textit{base type number}\\
& | \; \String & \textit{base type string}\\
& | \; ? & \textit{dynamic type}\\
& | \; T \rightarrow T & \textit{function types}\\
e ::= & & \textsc{expressions:}\\
& \;\; l & \textit{literals}\\
& | \; x & \textit{variables}\\
& | \; \lambda x{:}T{.}e & \textit{abstractions}\\
& | \; e_{1} e_{2} & \textit{application}
\end{array}
$$
\dend
\caption{Syntax of the gradually-typed lambda calculus}
\label{fig:gtlc}
\end{figure}

@@

Figure \ref{fig:consistent-subtyping} combines the consistency and subtyping
relations to compose the consistent-subtyping relation for the simply typed
lambda calculus extended with the dynamic type $?$.
When we combine consistency and subtyping, we are making subtyping handle
what casts are safe among static types, and we are making consistency
handle the casts that involve the dynamic type $?$.
The consistent-subtyping relation is not transitive, and thus
the dynamic type $?$ is not neutral to this relation.

@@

Another concern about casts is how to improve debugging support,
because a cast application can be delayed and the error related
to that cast application can appear considerable distance
from the real error.
\citet{wadler2009wpc} developed \emph{blame calculus} as a way
to handle this issue, and \citet{ahmed2011bfa} extended
blame calculus with polymorphism.
Blame calculus is an intermediate language to integrate
static and dynamic typing along with the blame tracking approach
proposed by \citet{findler2002chf}.

@@

There are some projects that incorporate gradual typing into some
programming languages.
Reticulated Python \citep{reticulated,vitousek2014deg} is a research
project that evaluates the costs of gradual typing in Python.
Gradualtalk \citep{allende2013gts} is a gradually-typed Smalltalk
that introduces a new cast insertion strategy for gradually-typed
objects \citep{allende2013cis}.
Grace \citep{black2012grace,black2013sg} is a new object-oriented,
gradually-typed, educational language.
In Grace, modules are gradually-typed objects, that is, modules
may have types and methods as attributes, and can have a state
\citep{homer2013modules}.
ActionScript \citep{moock2007as3} is one the first languages that
incorporated gradual typing to its implementation and
Perl 6 \citep{tang2007pri} is also being designed with gradual typing,
though there is few documentation about the gradual type systems
of these languages.

### 2015-03-15 doc/thesis/system.tex
Minor improvements to introduction, review and begining of typing rules

In this section we use a reduced core of Typed Lua to present the
typing rules that type check multiple assignment, table refinement,
and overloading on the return type of functions.
We limit this section to the rules that type check these features
because these rules are the most interesting rules of our type system.
Appendix \ref{app:rules} presents the full set of typing rules.

Our core limits control flow to if and while statements;
it has explicit type annotations, explicit scope for variables,
explicit method declarations, and explicit method calls.
Here is a list of features that are not present in our reduced core:
\begin{itemize}
\item labels and goto statements are not present because they
are difficult to handle along with our simplified form of
\emph{flow typing}, and they are out of scope for now;
\item explicit blocks are not present because we are already using
explicit scope for variables;
\item other loop structures such as repeat-until, numeric for,
and generic for are not present because we can use while to
express them;
\item fields in the table constructor are reduced to a single
form because we can use this form to express the missing forms; 
\item arithmetic operators are reduced to addition only
because the other operators have similar typing rules;
\item relational operators are reduced to equality and less than
because inequality has similar typing rules to equality and
other relational operators have similar typing rules to less than;
\item bitwise operators are reduced to bitwise and only
because the other operators have similar typing rules.
\end{itemize}

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\overline{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \; | \;
\mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s\\
& | \; \mathbf{return} \; el \; | \;
\lfloor a \rfloor_{0}\\
e ::= & \;\; \mathbf{nil} \; | \;
k \; | \;
id \; | \;
e_{1}[e_{2}] \; | \;
{<}T{>} \; id \; | \;
f \; | \;
\{ \; \overline{[e_{1}] = e_{2}} \; \} \; | \;
\{ \; \overline{[e_{1}] = e_{2}},me \; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \;\&\; e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
\# \; e \; | \;
\lfloor me \rfloor_{1}\\
l ::= & \;\; id_{l} \; | \;
e_{1}[e_{2}] \; | \;
id[k] \; {<}T{>}\\
k ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
el ::= & \;\; \overline{e} \; | \;
\overline{e}, me\\
me ::= & \;\; a \; | \;
{...}\\
a ::= & \;\; e(el) \; | \;
e{:}n(el)\\
m ::= & \;\; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el\\
f ::= & \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el\\
pl ::= & \;\; \overline{id{:}T} \; | \;
\overline{id{:}T},{...}{:}T\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax1}
\end{figure}

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rlr}
s ::= & & \textsc{statements:}\\
& \;\; \mathbf{skip} \;
| \; s_{1} \; ; \; s_{2} \;
| \; \overline{l} = el & \\
& | \; \mathbf{while} \; e \; \mathbf{do} \; s \;
| \; \mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2} & \\
& | \; \mathbf{local} \; \overline{id{:}T} = el \; \mathbf{in} \; s \;
| \; \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s & \\
& | \; \mathbf{rec} \; id{:}T = f \; \mathbf{in} \; s \;
| \; \mathbf{return} \; el \;
| \; \lfloor a \rfloor_{0} & \\
& | \; \mathbf{fun} \; id_{1}{:}id_{2} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
e ::= & & \textsc{expressions:}\\
& \;\; \mathbf{nil} \;
| \; k \;
| \; id \;
| \; e_{1}[e_{2}] \;
| \; {<}T{>} \; id \;
| \; f & \\
& | \; \{ \; \overline{[e_{1}] = e_{2}} \; \} \;
| \; \{ \; \overline{[e_{1}] = e_{2}},me \; \} & \\
& | \; e_{1} + e_{2} \;
| \; e_{1} \; {..} \; e_{2} \;
| \; e_{1} == e_{2} \;
| \; e_{1} < e_{2} & \\
& | \; e_{1} \;\&\; e_{2} \;
| \; e_{1} \; \mathbf{and} \; e_{2} \;
| \; e_{1} \; \mathbf{or} \; e_{2} \;
| \; \mathbf{not} \; e \;
| \; \# \; e & \\
& | \; \lfloor me \rfloor_{1} & \\
l ::= & & \textsc{left-hand values:}\\
& \;\; id_{l} \; | \;
e_{1}[e_{2}] \; | \;
id[k] \; {<}T{>} & \\
k ::= & & \textsc{literal constants:}\\
& \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} & \\
el ::= & & \textsc{expression lists:}\\
& \;\; \overline{e} \; | \;
\overline{e}, me & \\
me ::= & & \textsc{multiple results:}\\
& \;\; a \; | \;
{...} & \\
a ::= & & \textsc{applications:}\\
& \;\; e(el) \; | \;
e{:}n(el) & \\
f ::= & & \textsc{function declarations:}\\
& \;\; \mathbf{fun} \; (pl){:}S \; s \;;\; \mathbf{return} \; el & \\
pl ::= & & \textsc{parameter lists:}\\
& \;\; \overline{id{:}T} \; | \;
\overline{id{:}T},{...}{:}T & \\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax2}
\end{figure}

@@

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $k$ are literal constants, $p$ are table fields,
$me$ are expressions with multiple results, $f$ are function declarations,
$m$ are method declarations, $a$ are function and method applications,
$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$,
where $n$ can be zero.

@@

We present the typing rules as a deduction system for the typing relations
$\env_{1}, \penv \vdash s, \env_{2}$ and $\env_{1}, \penv \vdash e : T, \env_{2}$.
We use the first one for typing statements and the second one for typing expressions.
The first relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check a statement $s$
producing a new type environment $\env_{2}$.
The second relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check that an expression $e$ has
type $T$ and produces a new type environment $\env_{2}$.

Even though we can assign only first-level types to variables,
functions and methods can return unions of second-level types,
and our type system should be able to project these unions of
second-level types into unions of first-level types.
To do that, we use projection types and the environment $\penv$ that
maps projection variables to second-level types, while the
environment $\env$ maps variables to first-level types.
We will show how our type system uses the environment $\penv$ for
handling projection types in Section \ref{sec:fap}.
We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
with the variable $id$ that maps to type $T$.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$ with
the projection variable $x$ that maps to the second-level type $S$.

>>>>

In this section we use a reduced core of Typed Lua to present the
most interesting rules of our type system.
These rules type check multiple assignment, table refinement,
and overloading on the return type of functions.
Appendix \ref{app:rules} presents the full set of typing rules.

Our core limits control flow to if and while statements;
it has explicit type annotations, explicit scope for variables,
explicit method declarations, and explicit method calls.
Here is a list of features that are not present in our reduced core:
\begin{itemize}
\item labels and goto statements (they are difficult to handle along
with our simplified form of \emph{flow typing}, and they are out of
scope for now);
\item explicit blocks (we are already using explicit scope for variables);
\item other loop structures such as repeat-until, numeric for,
and generic for (we can use while to express them);
\item table fields other than $[e_{1}] = e_{2}$
(we can use this form to express the missing forms); 
\item arithmetic operators other than $+$
(other arithmetic operators have similar typing rules);
\item relational operators other than $==$ and $<$
(inequality has similar typing rules to $==$ and
other relational operators have similar typing rules to $<$);
\item bitwise operators other than $\&$
(other bitwise operators have similar typing rules).
\end{itemize}

Our reduced core does not lose much expressiveness, as it can express
any Lua program except those that use labels and goto statements.

@@

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$k$ are literal constants, $el$ are expression lists,
$me$ are expressions with multiple results, $a$ are function and method applications,
$f$ are function declarations, $pl$ are parameter lists,
$id$ are variable names, $T$ are first-level types, and $S$ are second-level types.
The notation $\overline{id{:}T}$ denotes the list $id_{1}{:}T_{1}, ..., id_{n}{:}T_{n}$,
where $n$ can be zero.

@@

Even though we can assign only first-level types to variables,
functions and methods can return unions of second-level types,
and our type system should be able to project these unions of
second-level types into unions of first-level types.
We use two different environments to handle this feature.
The first environment is the type environment $\env$ that maps
variables to first-level types.
We use $\env_{1}[id \mapsto T]$ to extend the environment $\env_{1}$
with the variable $id$ that maps to type $T$.
The second environment is the projection environment $\penv$ that
maps projection variables to second-level types.
We use $\penv[x \mapsto S]$ to extend the environment $\penv$
with the projection variable $x$ that maps to type $S$.
In Section \ref{sec:fap} we will show how our type system uses the
projection environment $\penv$ for handling projection types,
and also for projecting unions of second-level types into
unions of first-level types.

We present the typing rules as a deduction system for two typing relations,
one for typing statements and another for typing expressions.

We use the relation $\env_{1}, \penv \vdash s, \env_{2}$ for typing statements.
This relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check that a statement $s$
produces a new type environment $\env_{2}$.

We use the relation $\env_{1}, \penv \vdash e : T, \env_{2}$ for typing expressions.
This relation means that given a type environment $\env_{1}$
and a projection environment $\penv$, we can check that an expression $e$ has
type $T$ and produces a new type environment $\env_{2}$.
